Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2023-01-07 manuscript.tex
:pencil: feature refs added

Time series clustering has a number of conceptual use cases with psychological data. Prime among them is the ability to reduce the time, variable, and person complexity by extracting and organizing participant-level structures. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. These phenomena and principles can be meaningful differences that distinguish participants in different clusters, as well important patterns, trends, and relationships that participants share within a cluster \citep[e.g.,][]{schrodt2000}. Once distinct groups and patterns have been identified, researchers can examine the extent to which these within-group and between-group structures are associated with other variables of interest, such as personality traits, demographic characteristics, or other psychological constructs \citep[e.g.,][]{monden2022}. Once meaningful and robust structures and patterns are detected, time series clustering can, thus, be used to inform the development of robust theories as well as targeted interventions and therapies for individuals, for example, with mood disorders and other psychological conditions \citep[e.g.,][]{borsboom2021, eronen2020}.

@@

The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. In its most general approach, this can include any statistical summaries of the time series \citep[i.e., features; e.g.,][]{maharaj2019}. The extracted features can then be used as inputs to most clustering algorithms, including k-means or hierarchical clustering to group the participants into clusters \citep{wang2006}. The clustering of model parameters is, thus, technically a restrictive sub-form of feature-based clustering --- where lagged regression coefficients, for example, can be considered to be one type of feature capturing a time series. However, beyond model parameters, time series might also be described with any other statistical feature \citep{tiano2021}. Feature-based time series clustering has been a valued tool in many empirical fields for its flexibility and interpretability. As a result, features have been extracted for a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}.

With its rich history, a staggering variety of time series features have been proposed, not all of which are meaningful for psychological time series. As an example, \citet{wang2006} propose a set of nine statistical features for describing a trajectory: (1) trend, (2) seasonality, (3) periodicity, (4) skewness, (5) kurtosis, (6) serial correlation, (7) non-linearity, (8) self-similarity, and (9) chaos. This selection was also used by \citep{fulcher2013}. \citet{adya2001} proposed a total of 28 features relevant for forecasting, which largely fit within the selection proposed by \citet{wang2006}. A final illustration comes from a commonly used software package for feature extraction `tsfresh', which allows users to extract a total of 794 features to describe all kinds of aspects of a time series \citep[][]{christ2018}.

@@

\citet{kuppens2017} originally proposed (1) variability, (2) co-variance or ICC, (3) inertia, and (4) cross-lagged correlations. These features were then extended and formalized by \citet{krone2018} into (1) within-person variance, (2) innovation, (3) inertia or autocorrelation, (4) cross-lagged regression parameters, (5) covariances as granularity, and (6) mean intensity. \citet{krone2018} even built a parametric model to use a subset of these features to cluster study participants.

\citet{dejonckheere2019} discuss (1) mean, (2) variance, (3) relative variance, (4) instability, (5) inertia / autoregression, (6) interdependence / network density, (7) granularity or differentiation / ICC, (8) correlation, and (9) diversity / Gini coefficient.

Time series features have also come to include more summaries in terms of measurement distribution, categorical states, and, importantly, nonlinear developments. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data.

@@

\paragraph{Central Tendency} Familiar statistics from probability theory, the two features sit at the heart of many fundamental psychological questions. ``Are some people happier than others?'' (i.e., difference in central tendency). Mathematically, the central tendency has parametric and robust options to choose from (e.g., parametric: mean; robust: median).

\paragraph{Variability} "Are people settled in their attitudes towards migrants or do they fluctuate over time?'' (i.e., difference in variability). parametric: standard deviation, robust: median absolute deviation.

\paragraph{(In)stability} Stability captures the average change between two consecutive measurements. Do changes tend to be slow and gradual or fast and abrupt? Mathematically, (in)stability measurements often look at the average change between two consecutive measurements \citep[e.g., mean absolute change; e.g.,][]{barandas2020}.

\paragraph{Self-similarity} Self-similarity describes how much a measurement carries over to future measurements. There are two main ways in which experiences tend to be connected to future experiences --- resistance to change and seasonality. Both forms of self-similarity relate to different types of research questions. ``Do patients stay in a depressed mood for several measurements?'' (i.e., inertia), and ``Do participants drink more alcohol on Fridays?'' (i.e., periodicity). Both forms of self-similarity are often measured using autocorrelations with either a lag-1 (inertia) or the lag of the seasonality (e.g., seven days). Periodicity can additionally be captured using wave patterns \citep[e.g., fourier or wavelet transformation; e.g.,][]{mayor2022}.

>>>>

Time series clustering has a number of conceptual use cases with psychological data. Prime among them is the ability to reduce the time, variable, and person complexity by extracting and organizing participant-level structures. These reduction and structuring qualities can be essential in detecting phenomena and extracting more abstract functional principles \citep[][]{eronen2021a}. These phenomena and principles can be meaningful differences that distinguish participants in different clusters, as well important patterns, trends, and relationships that participants share within a cluster \citep[e.g.,][]{schrodt2000}. Once distinct groups and patterns have been identified, researchers can examine the extent to which these within-group and between-group structures are associated with other variables of interest, such as personality traits, demographic characteristics, or other psychological constructs \citep[e.g.,][]{monden2022}. By detecting meaningful and robust structures and patterns, time series clustering can, thus, be used to inform the development of robust theories as well as targeted interventions and therapies for individuals, for example, with mood disorders and other psychological conditions \citep[e.g.,][]{borsboom2021, eronen2020}.

@@

The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. In its most general approach, this can include any statistical summary of the time series \citep[i.e., feature; e.g.,][]{maharaj2019}. The extracted features can then be used as inputs to most clustering algorithms, including k-means or hierarchical clustering to group the participants into clusters \citep{wang2006}. The clustering of model parameters is, thus, technically a restrictive sub-form of feature-based clustering --- where lagged regression coefficients, for example, can be considered to be one type of feature capturing a time series. However, beyond model parameters, time series might also be described with any other statistical feature \citep{tiano2021}. Feature-based time series clustering has been a valued tool in many empirical fields for its flexibility and interpretability. As a result, features have been extracted for a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}.

With its rich history, a staggering variety of time series features have been proposed, not all of which are meaningful for psychological time series. As an example, \citet{wang2006} propose a set of nine statistical features for describing a trajectory: (1) trend, (2) seasonality, (3) periodicity, (4) skewness, (5) kurtosis, (6) serial correlation, (7) non-linearity, (8) self-similarity, and (9) chaos \citep[also see][]{fulcher2013}. \citet{adya2001} proposed a total of 28 features relevant for forecasting, which largely fit within the selection proposed by \citet{wang2006}. A final illustration comes from a commonly used software package for feature extraction `tsfresh', which allows users to extract a total of 794 features to describe all kinds of aspects of a time series \citep[][]{christ2018}.

@@

To understand emotion dynamics, \citet{kuppens2017} originally proposed four features: (1) within-person variability, (2) co-variance or ICC, (3) inertia or autocorrelation, and (4) cross-lagged correlations. These features were then extended by \citet{krone2018}, adding (5) innovation variance, and (6) mean intensity. \citet{krone2018} even built a parametric model to tentatively cluster study participants. From a slightly different perspective \citet{dejonckheere2019} later added three additional features for psychological time series: (7) instability (8) interdependence (i.e., network density), and (9) diversity (i.e., Gini coefficient)\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

@@

\paragraph{Central Tendency} Central tendency refers to the statistical measures that represent the ``middle" or ``average" of a set of data. The most common measures of central tendency are the mean, median, and mode (Hopkins, 2020). As a familiar statistic from probability theory, the central tendency sits at the heart of many fundamental questions about psychological time series. Researchers might, for example, be interested in whether ``Over a one month period, are some people happier than others?''. A recent review paper pointed out that the relatively simple mean statistic, is often overlooked in psychological time series analyses but is usually essential for understanding psychological differences \citep{bringmann2018c}. As such, several time series modeling approaches now advocate for conscious consideration of the mean \citep[e.g.,][]{bringmann2017a}. As with cross-sectional data, for time series data the appropriate measure of central tendency depends on the specific characteristics of the data being analyzed. For instance, if the data are normally distributed, the mean is often an appropriate measure. If the data are skewed or have outliers, the median may be a better choice \citep{weisberg1992}.

\paragraph{Variability} Variability, also referred to as dispersion or spread, captures to the degree to which a set of data differs from the central tendency \citep{weisberg1992}. In time series analyses, variability is important because it provides information about the distribution and diversity of the data. "Are people settled in their attitudes towards migrants or do they fluctuate over time?'' (i.e., difference in variability). 
There are several measures of variability, including the range, standard deviation, and variance (Hopkins, 2020). The appropriate measure to use depends on the specific characteristics of the data being analyzed. parametric: standard deviation, robust: median absolute deviation.

``Capturing variability may not just be helpful in identifying a worse psychological state or an inadequate emotion regulation strategy. It may also predict future course of symptoms. Affective variability at baseline predicted the recurrence of depressive symptoms over a three-year period in remitted depressed patients \citep{timm2017}. Similarly, elevated temporal auto-correlation and variance, as well as increased correlation between emotions, increased the probability for an upcoming shift between a normal and a depressed state \citep{vandeleemput2014}. ''\citep{myin-germeys2018}

EWS: anomalous variance \citep{wichers2019}

increase in variance \citep{scheffer2009}

\paragraph{(In)stability} Stability captures the average change between two consecutive measurements. While instability is conceptually related to the variability feature, variability does not take into account temporal dependency, whereas instability looks at the `jumpy-ness' of the data over time. In other words, instability reflects the fluctuation or inconsistency in a time series data over time, while variability reflects the range or diversity of values in a time series data \citep{trull2008}. For example, if a person has rapid and extreme changes in mood their mood is highly unstable, while if a person's mood responses span a wide range over the entire study period, their mood is highly variable \citep{jahng2008}. Common research questions with psychological time series might, thus, include: ``Do changes tend to be slow and gradual or fast and abrupt?" Mathematically, (in)stability measurements often look at the average change between two consecutive measurements \citep[e.g., mean absolute change; e.g.,][]{ebner-priemer2009,barandas2020}.

\paragraph{Self-similarity} Self-similarity describes how much a measurement carries over to future measurements. There are two main ways in which experiences tend to be connected to future experiences --- resistance to change and seasonality. Both forms of self-similarity relate to different types of research questions.

Resistance to change is the self-similarity of a measurement with its next measurement and is often referred to as inertia \citep{REF}. ``Do patients stay in a depressed mood for several measurements?'' 
characterize psychological maladjustment \citep{kuppens2010} and can even predict the initial onset of depression \citep{kuppens2012}.
Mathematically, inertia is often measured using autocorrelations or autoregressions with a lag-1.

Seasonality is the self-similarity of a measurement that returns in a cyclical manner and is often also referred to as periodicity \citep{REF}. ``Do participants drink more alcohol on Fridays?''. Both forms of self-similarity are often measured using autocorrelations with either a lag-1 (inertia) or the lag of the seasonality (e.g., seven days). Periodicity can additionally be captured using wave patterns \citep[e.g., fourier or wavelet transformation; e.g.,][]{mayor2022}.
